### 2016-10-03 Paper_v7.tex
update edits DB1

\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\vspace{0.05in}
{\par \centering Russell Gould$^{\ddag}$, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
\vspace{0.05in}
{\par \centering School of Chemical and Biomolecular Engineering and $^{\ddag}$Department of Biomedical Engineering, Cornell University, Ithaca NY 14853}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
\vspace{0.1in}
% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
%\vspace{0.1in}
{\par \centering \textbf{Character count:}~\emph{56,659}}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
\vspace{0.1in}
{\par \centering Jonathan T. Butcher,}
{\par \centering Associate Professor, Department of Biomedical Engineering,}
{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jtb47@cornell.edu}
{\par \centering Phone: (607) 255 - 3575}
{\par \centering Fax: (607) 255 - 7330}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
These models incorporate mass action kinetics within an ordinary differential equation (ODE) framework to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was maintained following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of vascular endothelial growth factor A (VEGF-A).
We incorporated mass action kinetics within an ordinary differential equation (ODE) framework to describe the EMT interaction network containing 995 gene, protein or mRNA components interconnected through 1700 interactions.
A family of model parameters was estimated using 41 molecular data sets generated in DLD1 colon carcinoma, MDCKII and A375 melanoma cells using the Pareto optimal ensemble technique (POETs) multiobjective optimization algorithm.
POETs identified more than 15,000 likely TGF-$\beta$ induced EMT models, from which we selected approximately 1100 models for further analysis.
Analysis of the model population suggested that both MCF10A and DLD1 cells could exhibit phenotypic heterogeneity if treated simultaneously with TGF-$\beta$1/2 and VEGF-A.
This heterogeneity was characterized by a significant fraction of the population being in a “hybrid state” having both high E-cadherin and high Vimentin expression.
We tested these predictions using qRT-PCR and flow-cytometry studies in a variety of experimental conditions.
Validation studies confirmed that upwards of 45\% of the cellular population could be put into the hybrid state in the presence of both TGF-$\beta$1/2 and VEGF-A.
Moreover, this response depended upon both activation of Sp1 by MAPK and NFATc1 transcriptional activity consistent with the predicted molecular signaling.
Lastly, the hybrid populations of both DLD1 and MCF10A cells exhibited different functional behavior than those from either TGF-$\beta$ or VEGF-A treatment.
The extent of ductal branch formation significantly increased with MCF10A cells in the hybrid phenotype, compared with cells treated with VEGF-A alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and reveal a novel signaling axis, which possibly regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

\subsection*{The model population captured key features of TGF-$\beta$ induced EMT}
The EMT model architecture, based upon curated molecular connectivity,
described the expression of 80 genes following exposure to TGF-$\beta$ isoforms and VEGF-A (Fig. \ref{fg:F1}).
The EMT model contained 995 molecular species interconnected by 1700 interactions.
Model equations were formulated using mass-action kinetics within an ordinary differential equation (ODE) framework.
ODEs and mass action kinetics are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
However, while ODE models can simulate complex intracellular behavior,
they require estimates for model parameters which are often difficult to obtain.
The EMT model had 1756 unknown model parameters, 1700 kinetic constants and 56 non-zero initial conditions.
As expected, these parameters were not uniquely identifiable given the training data \citep{Gadkar:2005vn}.
Thus, instead of identifying a single best fit (but uncertain) model,
we estimated a sloppy population of models (each consistent with the training data) by simultaneously minimizing the difference between model simulations and 41 molecular
data sets using the Pareto Optimal Ensemble Technique (POETs).
The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by POETs.
Additionally, we used 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
prediction error for each objective. Thus, we generated 11 different model ensembles.
Lastly, we compared model predictions with independent data sets not used during training (both at the molecular and model population
levels) to evaluate the predictive power of the parameter ensemble.
Additional details of the signaling architecture included in the model are presented in the materials and methods and the supplement.

POETs generated a population of probable signaling models which captured the multiple phases of EMT induction (Fig. \ref{fg:F2}).
POETs sampled well over 10$^6$ probable models during each stage of the cross-validation, using a combination of both local and global random sampling.
This sampling generated approximately 15,000 highly probable models from which we selected N $\simeq$ 1100 models for further analysis.
The selected models all had the same possible molecular connectivity, but different values for model parameters and extrinsic factors such as RNA polymerase or ribosome abundance.
Model selection was based upon Pareto rank, the prediction and training error across all objectives and model to model correlation (supplemental materials).
The model population recapitulated key signaling events following TGF-$\beta$ exposure.
We subdivided the response to TGF-$\beta$ exposure into two phases.
First, TGF-$\beta$1/2 signaling initiated a program which downregulated E-cadeherin expression in a MAPK dependent manner while
simultaneously upregulating TGF-$\beta$3 expression. Second, TGF$-\beta$3 secretion initiated an autocrine feedback which upregulated the expression of
mesenchymal markers such as Vimentin and key upstream transcription factors such as LEF-1 in a SMAD dependent manner.
Each phase involved the hierarchal expression and/or post-translational modification of several key transcription factors.
During the first phase, stimulation with TGF-$\beta$1/2 (10 a.u.) activated both the SMAD and MAPK pathways.
MAPK activation resulted in the phosphorylation of the transcription factor activator protein 1 (AP-1), which in-turn upregulated the expression of Snail, a well established transcriptional repressor (Fig. \ref{fg:F2}A).
Snail expression was MAPK-dependent; the MEK inhibitor U0126 blocked AP-1 activation and Snail expression following TGF-$\beta$1/2 exposure (Fig. \ref{fg:F2}A, Lane 3).
Similar results were obtained for Slug expression, confirming initial activation through the MAPK pathway (data not shown).
Overexpression of either Snail or Slug upregulated TGF-$\beta$3 expression (Fig. \ref{fg:F2}C) while simultaneously downregulating E-cadeherin expression (Fig. \ref{fg:F2}F).
During the second phase, TGF$-\beta$3 secretion and the subsequent autocrine signaling resulted in the upregulation of mesenchymal marker expression.
The TGF$-\beta$3 induced gene expression program involves a complex hierarchy of transcriptional and post-translational regulatory events.
Absence of E-cadherin indirectly promoted TGF$-\beta$3 expression through the $\beta$-catenin/TCF4 complex following Snail or Slug expression (Fig. \ref{fg:F2}C, Lane 2 or 3).
Conversely, over-expression of E-cadherin inhibited the TGF$-\beta$3 autocrine production by sequestering cytosolic $\beta$-catenin, thereby blocking EMT (Fig. \ref{fg:F2}C, Lane 4 or 5).
TGF$-\beta$3 signaled through through the Smad pathway to regulate LEF-1 expression and downstream target EMT genes (Fig. \ref{fg:F2}G).
TGF$-\beta$3 (10 a.u.) in combination with downstream inhibitors (DN-Smad4 and DN-LEF-1) completely inhibited Vimentin expression, while elevating E-cadherin expression (Fig. \ref{fg:F2}H,I).

The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
In whole, 78\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
generated from the best-fit nominal set (starting point for the optimization).
Conversely, we \emph{predicted} approximately 60\% of the training objectives, at a 95\% confidence interval compared to randomized parameters.
The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L).
This data was not used for model training.
The high predictability can be attributed to the combination of the leave-one-out cross validation scheme, diverse objective functions, and robustness of the POETs algorithm.
Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
A listing of data used for training is included in the supplement (Fig. \ref{fig:S5} and Fig. \ref{fig:S6}).

\subsection*{Identification of a novel LEF-1 regulator}
During model identification, we found that consistent TGF-$\beta$ induced EMT required an additional regulatory protein.
This protein, which we called hypothetical regulator 1 (YREG1), was required to mediate between SNAIL/SLUG transcriptional activity
and the upregulation of LEF-1 expression following TGF-$\beta$1/2 exposure.
SNAIL/SLUG are well known transcriptional repressors \citep{Hemavathy:2000aa,Hemavathy:2000ab,Dhasarathy:2011aa}, although there are a few studies
which suggest that at least SNAIL can also act as a transcriptional activator \citep{Guaita:2002aa}.
In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \citep{Jackstadt:2013aa}.
Thus, upon receiving a TGF-$\beta$1/2 signal, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \citep{Eastman:1999aa}.
Thus, either SNAIL/SLUG are acting as inducers (contrary to substantial biochemical evidence) or, they are repressing the expression of an intermediate repressor.
Given the biochemical evidence supporting SNAIL/SLUG as repressors, we created YREG1 a hypothetical intermediate repressor whose
expression is downregulated by SNAIL/SLUG. The literature data therefore suggested that YREG1 had two transcriptional targets, LEF-1 and TGF-$\beta$3.
By adding this regulator, our simulations became consistent with training and literature data.
Medici et al. suggested a similar idea where feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \citep{Medici:2008fk}.
Low levels of YREG1 expression were used in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
To test the potency of YREG1, we conducted knockdown and over-expression simulations following the addition of TGF-$\beta$1/2 (Fig. \ref{fig:S8}).
In the absence of YREG1, most of the population failed to consistently respond to TGF-$\beta$1/2 exposure compared to the wild-type (Fig. \ref{fig:S8}A).
Conversely, YREG1 overexpression revealed an exclusively epithelial phenotype following TGF-$\beta$1/2 stimulation (Fig. \ref{fig:S8}B).
Overexpression of YREG1 repressed LEF-1 and TGF-$\beta$3 expression,
thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex which promotes mesenchymal gene expression, or SMAD activity following from autocrine TGF-$\beta$3 signaling.
Likewise, the abundance of the pSmad2/4-LEF-1 complex was also reduced in cells overexpression YREG1, which blocked the repression of E-cadherin.
Taken together, we found that low YREG1 expression was necessary for stabilizing EMT,
while elevated YREG1 levels limited the extent of EMT induction.

@@

\subsection*{TGF$-\beta$1/2 and VEGF-A exposure promotes phenotype heterogeneity through NFATc and phosphorylated Sp1}
While we captured the central tendency of many of the molecular features of EMT induction following TGF$-\beta$1/2 exposure, an often neglected but important emergent
feature of developmental and pathological programs is population heterogeneity \citep{Park:2010fu}.
We (and others) previously hypothesized that deterministic model ensembles can interrogate population behavior, at least at a course grained level \citep{Lequieu:2011fj}.
We tested this hypothesis by analyzing the response of the population of EMT models to extracellular cues and then comparing this response to flow cytometry studies.
We used robustness coefficients to quantify the response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation.
We have previously used robustness coefficients to systematically quantify response of a system to structural or operational perturbations,
for example gene deletions or the addition of a growth factor or hormone  \citep{Song:2010fk,Tasseff:2010pi,Lequieu:2011fj,Tasseff:2011ve}.
Robustness coefficients quantify shifts in molecular marker abundance resulting from molecular of environmental perturbations relative to an unperturbed control state.
Robustness coefficients $\gg$ 1 indicate that marker abundance increased, while robustness coefficients $\ll$ 1 indicates marker abundance decreased relative to an unperturbed control.
A value of $\sim$ 1 indicates approximately no change in marker abundance following the perturbation.
We calculated robustness coefficients for each member of the ensemble (N $\simeq$ 1100) for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial)
following the addition of TGF$-\beta$1/2 alone (Fig. \ref{fg:F3}), and VEGF-A in combination with NFATc inhibitors (Fig. \ref{fg:F4}).
The absence of TGF$-\beta$1/2 or VEGF-A stimulation was used as the baseline for the robustness calculations.

@@

Our robustness analysis predicted that NFATc transcriptional activity was critical to maintaining E-cadherin expression in the presence of both VEGF-A and TGF$-\beta$1/2.
We experimentally tested this hypothesis by exposing both MCF10A (Fig. \ref{fg:F5}E,F) and DLD1 cells (Fig. \ref{fg:S4}) to combinations of VEGF-A and TGF$-\beta$1/2 in the presence or absence of VIVIT,
a soluble peptide inhibitor of NFATc transcriptional activity \citep{Aramburu:1999if}.
Treatment with VEGF-A (50ng/ml) and VIVIT (10$\mu$M) in MCF10A cells significantly reduced E-cadherin expression compared to VEGF-A alone (Fig 5D,E).
Co-treatment with VIVIT and TGF$-\beta$1/2 did not enhance EMT capacity of MCF10A cells above that of TGF$-\beta$1/2 alone (Fig 5A,B,E).
Likewise, VIVIT in combination with both TGF$-\beta$1/2 and VEGF-A resulted in a loss of E-cadherin gene and protein expression, while Slug and Vimentin levels remained increased (Fig. \ref{fg:F5}D,E ).
Quantitative flow cytometry confirmed these results in both MCF10A (Fig. \ref{fg:F5}F) and DLD1 cells (Fig. S4C).
Both epithelial cell lines initially had high levels of E-cadherin expression, and low vimentin abundance (Q1-99.5\%), but both MCF10A and DLD1 cells shifted from an epithelial to mesenchymal phenotype (Q1-33.4\%, Q4-42.8\%) following TGF$-\beta$1/2 exposure.
As expected, NFATc nuclear localization was repressed with VIVIT treatment regardless of ligand stimulation, while the abundance of nuclear phosphorylated Sp1 increased for both TGF$-\beta$1/2 and TGF$-\beta$1/2 + VIVIT conditions (Fig. 5D,E).
Combined TGF$-\beta$1/2 and VEGF-A increased both Vimentin and E-cadherin expression (Q1-42.1\%, Q2-52.3\%) compared to TGF$-\beta$1/2 alone.
Together, these results demonstrate that NFATc and phosphorylated Sp1 are critical for regulating E-cadherin and Vimentin expression during phenotype heterogeneity in MCF10A and DLD1.

>>>>

\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\vspace{0.05in}
{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
\vspace{0.05in}
{\par \centering School of Chemical and Biomolecular Engineering and $^{\ddag}$Department of Biomedical Engineering, Cornell University, Ithaca NY 14853}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
\vspace{0.1in}
% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
%\vspace{0.1in}
{\par \centering \textbf{Character count:}~\emph{56,659}}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
\vspace{0.1in}
{\par \centering Jonathan T. Butcher,}
{\par \centering Associate Professor, Department of Biomedical Engineering,}
{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jtb47@cornell.edu}
{\par \centering Phone: (607) 255 - 3575}
{\par \centering Fax: (607) 255 - 7330}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
These models incorporate mass action kinetics within an ordinary differential equation (ODE) framework to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was maintained following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of vascular endothelial growth factor A (VEGF-A).
We incorporated mass action kinetics within an ordinary differential equation (ODE) framework to describe the EMT interaction network containing 97 gene, protein or mRNA components interconnected through 169 interactions.
A family of model parameters was estimated using 41 molecular data sets generated in DLD1 colon carcinoma, MDCKII and A375 melanoma cells using the Pareto optimal ensemble technique (POETs) multiobjective optimization algorithm.
POETs generated an ensemble of approximately 1400 models for analysis.
Analysis of the model population suggested that both MCF10A and DLD1 cells could exhibit phenotypic heterogeneity if treated simultaneously with TGF-$\beta$1/2 and VEGF-A.
This heterogeneity was characterized by a significant fraction of the population being in a “hybrid state” having both high E-cadherin and high Vimentin expression.
We tested these predictions using qRT-PCR and flow-cytometry studies in a variety of experimental conditions.
Validation studies confirmed that upwards of 45\% of the cellular population could be put into the hybrid state in the presence of both TGF-$\beta$1/2 and VEGF-A.
Moreover, this response depended upon both activation of Sp1 by MAPK and NFATc1 transcriptional activity consistent with the predicted molecular signaling.
Lastly, the hybrid populations of both DLD1 and MCF10A cells exhibited different functional behavior than those from either TGF-$\beta$ or VEGF-A treatment.
The extent of ductal branch formation significantly increased with MCF10A cells in the hybrid phenotype, compared with cells treated with VEGF-A alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and reveal a novel signaling axis, which possibly regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

\subsection*{The model population captured key features of TGF-$\beta$ induced EMT}
The EMT model architecture, based upon curated molecular connectivity,
described the expression of 23 genes following exposure to TGF-$\beta$ isoforms and VEGF-A (Fig. \ref{fg:F1}).
The EMT model contained 74 molecular species interconnected by 169 interactions.
Model equations were formulated using mass-action kinetics within an ordinary differential equation (ODE) framework.
ODEs and mass action kinetics are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
However, while ODE models can simulate complex intracellular behavior,
they require estimates for model parameters which are often difficult to obtain.
The EMT model had 251 unknown model parameters, 169 kinetic constants and 45 non-zero initial conditions.
As expected, these parameters were not uniquely identifiable given the training data \citep{Gadkar:2005vn}.
Thus, instead of identifying a single best fit (but uncertain) model,
we estimated a sloppy population of models (each consistent with the training data) by simultaneously minimizing the difference between model simulations and 41 molecular
data sets using the Pareto Optimal Ensemble Technique (POETs).
The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by POETs.
Additionally, we used 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
prediction error for each objective. Thus, we generated 11 different model ensembles.
Lastly, we compared model predictions with independent data sets not used during training (both at the molecular and model population
levels) to evaluate the predictive power of the parameter ensemble.
Additional details of the signaling architecture included in the model are presented in the materials and methods and the supplement.

POETs generated a population of probable signaling models which captured the multiple phases of EMT induction (Fig. \ref{fg:F2}).
POETs sampled well over 10$^4$ probable models during each stage of the cross-validation #####, using a combination of both local and global random sampling.
From this analysis, N $\simeq$ 1400 models were selected for further analysis.
The selected models all had the same possible molecular connectivity, but different values for model parameters. Trascription and translation rates, as well as mRNA and protein degradation terms, were set using physical values from the literature, and allowed to vary by a scaling factor, see methods ####citebionumbers####.
Model selection was based upon Pareto rank, the prediction and training error across all objectives and model to model correlation (supplemental materials).
The model population recapitulated key signaling events following TGF-$\beta$ exposure.
We subdivided the response to TGF-$\beta$ exposure into two phases.
First, TGF-$\beta$1/2 signaling initiated a program which downregulated E-cadeherin expression in a MAPK dependent manner while
simultaneously upregulating TGF-$\beta$3 expression. Second, TGF$-\beta$3 secretion initiated an autocrine feedback which upregulated the expression of
mesenchymal markers such as Vimentin and key upstream transcription factors such as LEF-1 in a SMAD dependent manner.
Each phase involved the hierarchal expression and/or post-translational modification of several key transcription factors.
During the first phase, stimulation with TGF-$\beta$1/2 (10 a.u.) activated both the SMAD and MAPK pathways.
MAPK activation resulted in the phosphorylation of the transcription factor activator protein 1 (AP-1), which in-turn upregulated the expression of Snail, a well established transcriptional repressor (Fig. \ref{fg:F2}A).
Snail expression was MAPK-dependent; the MEK inhibitor U0126 blocked AP-1 activation and Snail expression following TGF-$\beta$1/2 exposure (Fig. \ref{fg:F2}A, Lane 3).
Similar results were obtained for Slug expression, confirming initial activation through the MAPK pathway (data not shown).
Overexpression of either Snail or Slug upregulated TGF-$\beta$3 expression (Fig. \ref{fg:F2}C) while simultaneously downregulating E-cadeherin expression (Fig. \ref{fg:F2}F).
During the second phase, TGF$-\beta$3 secretion and the subsequent autocrine signaling resulted in the upregulation of mesenchymal marker expression.
The TGF$-\beta$3 induced gene expression program involves a complex hierarchy of transcriptional and post-translational regulatory events.
Absence of E-cadherin indirectly promoted TGF$-\beta$3 expression through the $\beta$-catenin/TCF4 complex following Snail or Slug expression (Fig. \ref{fg:F2}C, Lane 2 or 3).
Conversely, over-expression of E-cadherin inhibited the TGF$-\beta$3 autocrine production by sequestering cytosolic $\beta$-catenin, thereby blocking EMT (Fig. \ref{fg:F2}C, Lane 4 or 5).
TGF$-\beta$3 signaled through the Smad pathway to regulate LEF-1 expression and downstream target EMT genes (Fig. \ref{fg:F2}G).
TGF$-\beta$3 (10 a.u.) in combination with downstream inhibitors (DN-Smad4 and DN-LEF-1) completely inhibited Vimentin expression, while elevating E-cadherin expression (Fig. \ref{fg:F2}H,I).

The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
In whole, 72\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
generated from a random starting point.
Conversely, we \emph{predicted} approximately 72\% of the training objectives, at a 95\% confidence interval compared to randomized parameters.
The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L).
This data was not used for model training.
The high predictability can be attributed to the combination of the leave-one-out cross validation scheme, diverse objective functions, and robustness of the POETs algorithm.
Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
A listing of data used for training is included in the supplement (Fig. \ref{fig:S5} and Fig. \ref{fig:S6}).

\subsection*{Identification of a novel LEF-1 regulator}
During model identification, we found that consistent TGF-$\beta$ induced EMT required an additional regulatory protein.
This protein, which we called hypothetical regulator 1 (YREG1), was required to mediate between SNAIL/SLUG transcriptional activity
and the upregulation of LEF-1 expression following TGF-$\beta$1/2 exposure.
SNAIL/SLUG are well known transcriptional repressors \citep{Hemavathy:2000aa,Hemavathy:2000ab,Dhasarathy:2011aa}, although there are a few studies
which suggest that at least SNAIL can also act as a transcriptional activator \citep{Guaita:2002aa}.
In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \citep{Jackstadt:2013aa}.
Thus, upon receiving a TGF-$\beta$1/2 signal, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \citep{Eastman:1999aa}.
Thus, either SNAIL/SLUG are acting as inducers (contrary to substantial biochemical evidence) or, they are repressing the expression of an intermediate repressor.
Given the biochemical evidence supporting SNAIL/SLUG as repressors, we created YREG1 a hypothetical intermediate repressor whose
expression is downregulated by SNAIL/SLUG. The literature data therefore suggested that YREG1 had two transcriptional targets, LEF-1 and TGF-$\beta$3.
By adding this regulator, our simulations became consistent with training and literature data.
Medici et al. suggested a similar idea where feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \citep{Medici:2008fk}.
Low levels of YREG1 expression were present in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
To test the potency of YREG1, we conducted knockdown and over-expression simulations following the addition of TGF-$\beta$1/2 (Fig. \ref{fig:S8}).
In the absence of YREG1, most of the population failed to consistently respond to TGF-$\beta$1/2 exposure compared to the wild-type (Fig. \ref{fig:S8}A).
Conversely, YREG1 overexpression revealed an exclusively epithelial phenotype following TGF-$\beta$1/2 stimulation (Fig. \ref{fig:S8}B).
Overexpression of YREG1 repressed LEF-1 and TGF-$\beta$3 expression,
thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex which promotes mesenchymal gene expression, or SMAD activity following from autocrine TGF-$\beta$3 signaling.
Taken together, we found that low YREG1 expression was necessary for stabilizing EMT,
while elevated YREG1 levels limited the extent of EMT induction.

@@

\subsection*{TGF$-\beta$1/2 and VEGF-A exposure promotes phenotype heterogeneity through NFATc and phosphorylated Sp1}
While we captured the central tendency of many of the molecular features of EMT induction following TGF$-\beta$1/2 exposure, an often neglected but important emergent
feature of developmental and pathological programs is population heterogeneity \citep{Park:2010fu}.
We (and others) previously hypothesized that deterministic model ensembles can interrogate population behavior, at least at a course grained level \citep{Lequieu:2011fj}.
We tested this hypothesis by analyzing the response of the population of EMT models to extracellular cues and then comparing this response to flow cytometry studies.
We quantiied the phenotypic response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation as a function of physical measurements of abundance of markers in system.
We evlautated these quantities for each member of the ensemble (N $\simeq$ 1400) for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial)
following the addition of TGF$-\beta$1/2 alone (Fig. \ref{fg:F3}), and VEGF-A in combination with NFATc inhibitors (Fig. \ref{fg:F4}).
We were further able to use this analysis to examine the activity and effects of underlying signaling molecules.

@@

Our phenotypic analysis predicted that NFATc transcriptional activity was critical to maintaining E-cadherin expression in the presence of both VEGF-A and TGF$-\beta$1/2.
We experimentally tested this hypothesis by exposing both MCF10A (Fig. \ref{fg:F5}E,F) and DLD1 cells (Fig. \ref{fg:S4}) to combinations of VEGF-A and TGF$-\beta$1/2 in the presence or absence of VIVIT,
a soluble peptide inhibitor of NFATc transcriptional activity \citep{Aramburu:1999if}.
Treatment with VEGF-A (50ng/ml) and VIVIT (10$\mu$M) in MCF10A cells significantly reduced E-cadherin expression compared to VEGF-A alone (Fig 5D,E).
Co-treatment with VIVIT and TGF$-\beta$1/2 did not enhance EMT capacity of MCF10A cells above that of TGF$-\beta$1/2 alone (Fig 5A,B,E).
Likewise, VIVIT in combination with both TGF$-\beta$1/2 and VEGF-A resulted in a loss of E-cadherin gene and protein expression, while Slug and Vimentin levels remained increased (Fig. \ref{fg:F5}D,E ).
Quantitative flow cytometry confirmed these results in both MCF10A (Fig. \ref{fg:F5}F) and DLD1 cells (Fig. S4C).
Both epithelial cell lines initially had high levels of E-cadherin expression, and low vimentin abundance (Q1-99.5\%), but both MCF10A and DLD1 cells shifted from an epithelial to mesenchymal phenotype (Q1-33.4\%, Q4-42.8\%) following TGF$-\beta$1/2 exposure.
As expected, NFATc nuclear localization was repressed with VIVIT treatment regardless of ligand stimulation, while the abundance of nuclear phosphorylated Sp1 increased for both TGF$-\beta$1/2 and TGF$-\beta$1/2 + VIVIT conditions (Fig. 5D,E).
Combined TGF$-\beta$1/2 and VEGF-A increased both Vimentin and E-cadherin expression (Q1-42.1\%, Q2-52.3\%) compared to TGF$-\beta$1/2 alone.
Together, these results demonstrate that NFATc and phosphorylated Sp1 are critical for regulating E-cadherin and Vimentin expression during phenotype heterogeneity in MCF10A and DLD1.
